---
title: "ProBatchFeatures"
author: |
  | Yuliya Burankova
  | Institute for Computational Systems Biology, University of Hamburg, Germany
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
bibliography: "references.bib"
csl: "nature-no-superscript.csl"
package: proBatch
vignette: >
  %\VignetteIndexEntry{ProBatchFeatures: QFeatures-based pipelines with operation logging for proBatch}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
abstract: |
  This vignette demonstrates how to perform and evaluate data preprocessing and batch-effect correction workflows using `ProBatchFeatures`, a `QFeatures`-based extension of `proBatch`.
editor_options:
  markdown:
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, warning = FALSE, message = FALSE, fig.pos = "h",
    collapse = TRUE,
    comment = "#>",
    knitr.table.format = "simple"
)
```

```{r setup, include = FALSE}
options(tinytex.verbose = TRUE)
```

# Overview

`ProBatchFeatures` is an extension of `proBatch` based on `QFeatures`
[@QFeatures], which extends its functionality by storing each processing stage
(for example, raw data → log-transformation → normalization → batch effect
correction → …) as a new `SummarizedExperiment` assay, and logging every step.
It maintains full compatibility with `QFeatures` functions, and facilitates
quick benchmarking of data preprocessing pipelines.

This vignette demonstrates how to:

1. Construct a `ProBatchFeatures` object from
    matrix-formatted quantification data.

2. Apply and record standard preprocessing steps (filtering features with too
    many missing values, log-transformation, normalization, and batch-effect
    correction) in a chained workflow.

3. Perform diagnostic checks, including PCA, t-SNE, and UMAP.

4. Retrieve processed assay data and the corresponding preprocessing pipeline
    for benchmarking or reuse.


# Setup


## Installation

To install the latest version of `proBatch` package, use `BiocManager` package:

```{r install_proBatch, fig.show='hold', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("proBatch")
```

## Loading required packages

This vignette uses `dplyr`, `tibble`, and `ggplot2` (alongside `proBatch`) for
data manipulation and visualization.

```{r load_packages, message=FALSE}
library(proBatch)

library(dplyr)
library(tibble)
library(ggplot2)
library(QFeatures)
```


## Loading and preparing the example dataset

As an example, we will use the *E. coli* dataset described in the FedProt paper
[@burankova2025privacy]:

```{r load_data}
data("example_ecoli_data", package = "proBatch")
```

The dataset consists of data from five centers that were  measured
and analyzed independently.
For simplicity, data from all centers have been merged into a single dataset,
and a subset containing 400 protein groups and their corresponding peptides
is used in this example.

```{r extract-data}
# Extracting data
all_metadata <- example_ecoli_data$all_metadata
all_precursors <- example_ecoli_data$all_precursors
all_protein_groups <- example_ecoli_data$all_protein_groups
all_precursor_pg_match <- example_ecoli_data$all_precursor_pg_match

# Keeping only essential
rm(example_ecoli_data)
```

# Constructing a `ProBatchFeatures` object

The dataset contains two hierarchical levels. While additional levels can be
added in a manner similar to `QFeatures`, this example focuses on the peptide
and protein group levels.
The `all_precursor_pg_match` dataframe stores the mapping between these levels.
We begin by using the  `ProBatchFeatures()` constructor to create an object
representing the peptide-level data:

```{r build-pbf-from-matrix}
# Building from a (features x samples) matrix
pbf <- ProBatchFeatures(
    data_matrix = all_precursors, # matrix of peptide intensities (features x samples)
    sample_annotation = all_metadata, # data.frame of sample metadata
    sample_id_col = "Run", # column in metadata that matches sample IDs
    level = "peptide" # label this assay level as "peptide"
)

pbf # show basic info about the ProBatchFeatures object
```

Assay names follow the convention "<level>::<pipeline>", for
example, "peptide::raw", or
"protein::median_on_log".

Next, we add the protein-level data as a second assay in the same
`ProBatchFeatures` instance.
Internally, this is achieved by creating a `SummarizedExperiment` for
the protein-level data and
adding it to the existing object using `QFeatures` functionality.
The same sample annotation should be used for the new assay to ensure
correct column alignment:

```{r add-new-pbf-level}
# Adding proteins as a new level and mapping them to precursors
#    all_precursor_pg_match has columns: "Precursor.Id", "Protein.Ids"
pbf <- pb_add_level(
    object = pbf,
    from = "peptide::raw",
    new_matrix = all_protein_groups,
    to_level = "protein", # will name "protein::raw" by default
    mapping_df = all_precursor_pg_match,
    from_id = "Precursor.Id",
    to_id = "Protein.Ids",
    map_strategy = "as_is"
)

pbf
```

If a precursor (peptide) maps to multiple protein groups, `map_strategy`
specifies how to resolve multimappers. Available options are `"first"`,
`"longest"`, and `"as_is"`. The first two strategies select a single
ProteinID per peptide: `"first"` takes the first match in the mapping
data frame, while `"longest"` prefers the match with the most
semicolon-separated identifiers, then the longest string, then the first
occurrence.
`"as_is"` requires a one-to-one mapping between identifiers.


```{r check_pbf}
# Check
validObject(pbf) # should be TRUE
assayLink(pbf, "protein::raw") # verify link summary

# Remove unused variables to clean up the workspace
rm(all_metadata, all_precursor_pg_match, all_precursors, all_protein_groups)
```

# Processing pipeline with diagnostics

With the data now loaded into a `ProBatchFeatures` object with peptide- and
protein-level assays, we can demonstrate an example of typical preprocessing
workflow. This pipeline includes filtering low-quality features, applying a
log2-transformation,  median normalization, and batch effect correction.


## Step 1 - filtering features with too many missing values

Features detected in only a small subset of samples are unreliable and may
obscure systematic effects in subsequent analyses. To examine the proportion of
missing values across datasets, `pb_nNA()` function is used:

Function `pb_nNA()` summarizes missingness per assay:

```{r nNA_calc}
# pb_nNA returns a DataFrame with counts of NAs per feature/sample
na_counts <- pb_nNA(pbf)

# na_counts contains info about total number of NAs and % in the data:
na_counts[["nNA"]] %>% as_tibble()
```
Here nNA is a number of missing values in the assay and pNA is its
proportion. It is possible to inspect the per-feature and per-sample
breakdown for each assay:

```{r nNA_individual}
# check NAs per sample:
head(na_counts[["peptide::raw"]]$nNAcols) %>% as_tibble()
```

For each assay the following tables are available:
```{r print_na_counts}
print(names(na_counts[["peptide::raw"]]))
```

Visualisations help explore batch-specific patterns of missingness. Firstly, we
define a color scheme which will be used in all diagnostic plots:

```{r define_colors}
# color scheme for lab and condition labels
color_scheme <- sample_annotation_to_colors(
    pbf,
    sample_id_col = "Run",
    factor_columns = c("Lab", "Condition"),
    numeric_columns = NULL
)
```

Displaying features that contain at least one missing as a heatmap: 

```{r plot_na_heatmap_single, fig.show='hold', fig.width=10, fig.height=4}
plot_NA_heatmap(
    pbf, # by default the last created assay
    show_rownames = FALSE, show_row_dend = FALSE,
    color_by = "Lab"
)
```

To display all features, set `drop_complete = FALSE`. When the number of
features or samples exceeds 5000, a random subset of 5000 rows or columns
is used for visualization by default. To display the entire dataset, set
`use_subset = FALSE`.

To display multiple assays in one plot,  specify them
using `pbf_name` parameter:

```{r plot_na_heatmap_multi, fig.show='hold', fig.width=10, fig.height=4}
plot_NA_heatmap(
    pbf,
    pbf_name = c("peptide::raw", "protein::raw"),
    show_rownames = FALSE, show_row_dend = FALSE,
    color_by = "Lab"
)
```

The overlap between peptide and protein identifications across samples can
be visualized as follows:

```{r plot_na_frequency_panel, fig.show='hold', fig.width=7, fig.height=4}
plot_NA_frequency(
    pbf,
    pbf_name = c("peptide::raw", "protein::raw"),
    show_percent = FALSE
)
```

In the resulting plot, particularly for peptides, batch-specific patterns are
evident, reflecting the 19-20 samples contributed by each laboratory.

Here, we apply a simple filter that retains only features quantified in at
least 40% of the samples.
This threshold can be adjusted if necessary.

```{r apply_missing_filter}
pbf <- pb_filterNA(
    pbf, # without specifying, the filter will be applied to all assays in the object
    inplace = TRUE, # if false, filtered matrix will be saved as a new assay.
    pNA = 0.6 # the maximal proportion of missing values per feature
)
pbf
```

After filtering, the `ProBatchFeatures` object stores fewer features,
which sharpens downstream diagnostics.

```{r plot_na_heatmap_filtered, fig.show='hold', fig.width=7, fig.height=4}
plot_NA_heatmap(
    pbf,
    pbf_name = c("peptide::raw", "protein::raw"),
    show_rownames = FALSE, show_row_dend = FALSE,
    color_by = "Lab", # currently only one level is supported
    border_color = NA # pheatmap parameter to remove cell borders
)
```

However, as the plots indicate, the remaining patterns of missingness are still associated with the lab.  
To remove the residual systematic effects from the data, we further perform
normalization and batch-effect correction.

## Step 2 - log2-transformation

Proteomics data are commonly log-transformed to stabilize variance for
statistical analysis and data visualization. Here, we perform a
log2-transformation of both peptide- and protein-level data:

```{r run_log_transform}
pbf <- log_transform_dm(
    pbf,
    log_base = 2, offset = 1,
    pbf_name = "protein::raw"
)

pbf <- log_transform_dm(
    pbf,
    log_base = 2, offset = 1,
    pbf_name = "peptide::raw"
)

pbf
```

After log-transformation, `pbf` records the virtual targets
`peptide::log2_on_raw` and `protein::log2_on_raw`. Because log-transformation is
fast to recompute, these results are replayed when requested instead of being
stored as assays by default. Set `store_fast_steps = TRUE` or supply an
explicit `final_name` to materialize them.


The intensity distributions of each sample after log-transformation can be
visualized as boxplots to identify potential batch-specific shifts:

```{r plot_logmean, fig.height=6, fig.width=8, warning=FALSE}
plot_boxplot(
    pbf,
    pbf_name = c(
        "peptide::log2_on_raw", "protein::log2_on_raw"), # plot multiple on one
    sample_id_col = "Run",
    order_col = "Run",
    batch_col = "Lab",
    color_by_batch = TRUE,
    color_scheme = color_scheme,
    base_size = 7,
    plot_ncol = 1, # how many columns use for plotting multy-assay plot
    plot_title = c( # title for each assay
        "Peptide level - log2 scale",
        "Protein level - log2 scale"
    )
)
```

Additionally, the intensity distribution of peptides and
proteins with and without NAs
can be plotted:

```{r plot_na_density_panels, fig.show='hold', fig.width=10, fig.height=4}
plot_NA_density(
    pbf,
    pbf_name = c("peptide::log2_on_raw", "protein::log2_on_raw")
)
```
The comparison confirms that log2 stabilises the variance for both
peptides and proteins, yet the cross-lab shifts in median intensity
persist and will need to be addressed downstream.


Next, we will perform a Principal Component Analysis (PCA) to get a
high-level overview of the sample clustering. We expect to see a strong
batch effect, where samples cluster by their center of origin ("Lab")
rather than their biological condition ("Condition").

```{r plot_PCA, fig.show='hold', fig.width=8, fig.height=4}
pca1 <- plot_PCA(
    pbf,
    pbf_name = "protein::log2_on_raw",
    sample_id_col = "Run",
    color_scheme = color_scheme,
    color_by = "Lab",
    shape_by = "Condition",
    fill_the_missing = NULL, # remove incomplete feature rows for this plot
    plot_title = "NA rows removed, protein, log2",
    base_size = 10, point_size = 3, point_alpha = 0.5
)

pca2 <- plot_PCA(
    pbf,
    pbf_name = "protein::log2_on_raw",
    sample_id_col = "Run",
    color_scheme = color_scheme,
    color_by = "Condition",
    shape_by = "Lab",
    fill_the_missing = 0, # default value is -1
    plot_title = "NA replaced with 0, protein, log2",
    base_size = 10, point_size = 3, point_alpha = 0.5
)

library(gridExtra)
grid.arrange(pca1, pca2, ncol = 2, nrow = 1)
```

Similarly, the PCA plots clearly demonstrate that the primary source of
variance in the raw data is the lab.

## Step 3 - median normalization

To make the samples more comparable by removing the effect of sample load on
intensities, we apply median
normalization:

```{r apply_median_normalization}
pbf <- pb_transform(
    object = pbf,
    from = "peptide::log2_on_raw",
    steps = "medianNorm"
)

pbf <- pb_transform(
    object = pbf,
    from = "protein::log2_on_raw",
    steps = "medianNorm"
)

pbf
```

`pb_transform()` appends `peptide::medianNorm_on_log2_on_raw` and
`protein::medianNorm_on_log2_on_raw`. We can now plot and compare these assays
before and after normalization:

```{r compare_median_norm_diagnostics, fig.show='hold', fig.width=8, fig.height=4}
# boxplot
plot_boxplot(
    pbf,
    sample_id_col = "Run",
    order_col = "Run",
    batch_col = "Lab",
    color_by_batch = TRUE,
    color_scheme = color_scheme,
    base_size = 7,
    pbf_name = c("protein::log2_on_raw", "protein::medianNorm_on_log2_on_raw"),
    plot_ncol = 1,
    plot_title = c(
        "Before Median Normalization, protein level",
        "After Median Normalization, protein level"
    )
)

# plot PCA plot
plot_PCA(
    pbf,
    pbf_name = c("protein::log2_on_raw", "protein::medianNorm_on_log2_on_raw"),
    sample_id_col = "Run",
    color_scheme = color_scheme,
    color_by = "Lab",
    shape_by = "Condition",
    fill_the_missing = NULL, # remove incomplete feature rows for these plots
    plot_title = c(
        "Before Median Normalization, protein level",
        "After Median Normalization, protein level"
    ),
    base_size = 10, point_size = 3, point_alpha = 0.5
)
```

Although median normalization made intensity distributions more similar, the
residual lab-specific batch effects remain evident on the PCA plots
and need to be corrected.

## Step 4 - batch-effect correction

The user can choose from two Core-registered batch-effect correction
steps: `combat`, backed by `ComBat` [@Johnson2006] from **sva**, and
`limmaRBE`, backed by `removeBatchEffect()` from **limma**
[@Ritchie2015]. Both steps can be applied and logged with
`pb_transform()`.

Correction methods use the common missing-value policies `"error"`,
`"keep"`, `"drop_features"`, and `"fill"`. The default, `"error"`,
stops when values are missing. `"keep"` passes missing values to a
method that supports them, `"drop_features"` removes incomplete feature
rows, and `"fill"` requires an explicit numeric `fill_value`. Here we
choose `"drop_features"` so both correction backends receive complete
feature rows.

```{r prepare_sample_annotations}
# sample annotations used by both correction methods
sa <- colData(pbf) %>% as.data.frame()
```

Apply **limma**'s `removeBatchEffect()` to log-transformed and normalized data:

```{r apply_limma_correction}
# limma::removeBatchEffect method
pbf <- pb_transform(
    pbf,
    from = "protein::log2_on_raw",
    steps = "limmaRBE",
    params_list = list(list(
        sample_annotation = sa,
        batch_col = "Lab",
        covariates_cols = c("Condition"),
        fill_the_missing = "drop_features"
    ))
)

pbf <- pb_transform(
    pbf,
    from = "protein::medianNorm_on_log2_on_raw",
    steps = "limmaRBE",
    params_list = list(list(
        sample_annotation = sa,
        batch_col = "Lab",
        covariates_cols = c("Condition"),
        fill_the_missing = "drop_features"
    ))
)
```

Because batch-effect correction is not a `fast step`, two new
protein-level assays are materialized:

- `protein::limmaRBE_on_log2_on_raw`
- `protein::limmaRBE_on_medianNorm_on_log2_on_raw`

Both contain only complete feature rows because this example requested
`"drop_features"`.

```{r inspect_pbf_after_limma}
print(pbf)
```

Alternatively, batch effect correction can be performed using `ComBat`:

```{r apply_combat_correction}
# sva::ComBat wrapped via pb_transform()
pbf <- pb_transform(
    pbf,
    from = "protein::log2_on_raw",
    steps = "combat",
    params_list = list(list(
        sample_annotation = sa,
        batch_col = "Lab",
        sample_id_col = "Run",
        par.prior = TRUE,
        fill_the_missing = "drop_features"
    ))
)

pbf <- pb_transform(
    pbf,
    from = "protein::medianNorm_on_log2_on_raw",
    steps = "combat",
    params_list = list(list(
        sample_annotation = sa,
        batch_col = "Lab",
        sample_id_col = "Run",
        par.prior = TRUE,
        fill_the_missing = "drop_features"
    ))
)
```
The `ComBat` method from the **sva** package cannot be applied to data
containing missing values. We therefore requested
`fill_the_missing = "drop_features"` explicitly. To replace missing
values with a constant instead, use `fill_the_missing = "fill"` together
with a finite numeric `fill_value`; omitting a policy retains the safer
`"error"` default.

```{r inspect_pbf_after_combat}
print(pbf)
```

Similarly, two new protein-level ComBat assays are materialized:

- `protein::combat_on_log2_on_raw`
- `protein::combat_on_medianNorm_on_log2_on_raw`

The corrected assays contain fewer protein groups than the filtered raw
data whenever incomplete features were dropped.

All steps can be applied to peptide-level data in the same way:

```{r apply_batch_correction_peptide}
pbf <- pb_transform(
    pbf,
    from = "peptide::medianNorm_on_log2_on_raw",
    steps = "limmaRBE",
    params_list = list(
        list(
            sample_annotation = sa,
            batch_col = "Lab",
            covariates_cols = c("Condition"),
            fill_the_missing = "drop_features"
        )
    )
)

pbf <- pb_transform(
    pbf,
    from = "peptide::medianNorm_on_log2_on_raw",
    steps = "combat",
    params_list = list(
        list(
            sample_annotation = sa,
            batch_col = "Lab",
            sample_id_col = "Run",
            par.prior = TRUE,
            fill_the_missing = "drop_features"
        )
    )
)

pbf
```

Two new peptide-level assays are materialized:

- `peptide::limmaRBE_on_medianNorm_on_log2_on_raw`
- `peptide::combat_on_medianNorm_on_log2_on_raw`

As at the protein level, each contains only feature rows that were
complete across samples under the requested policy.

## Step 5 - assess processed assays

The four new assays (`protein::limmaRBE_on_…` and `protein::combat_on_…`) are
now stored alongside the previous processing steps. We can explore these assays
using PCA to verify that the batch effect has been successfully removed and
compare them to the data before batch effect correction.

### Principal component analysis

The progression from log$_2$-transformed to batch-corrected assays demonstrates
the expected reduction in lab-specific variance and a clearer separation by
biological condition in the final panel.

```{r plot_pca_limma_series, fig.height=7, fig.show='hold', fig.width=8, message=FALSE, warning=FALSE}
plot_PCA(
    pbf,
    pbf_name = c(
        "protein::log2_on_raw",
        "protein::medianNorm_on_log2_on_raw",
        "protein::limmaRBE_on_medianNorm_on_log2_on_raw",
        "protein::combat_on_medianNorm_on_log2_on_raw"
    ),
    sample_id_col = "Run",
    color_scheme = color_scheme,
    color_by = "Lab",
    shape_by = "Condition",
    fill_the_missing = NULL,
    base_size = 8, point_size = 3, point_alpha = 0.5
)
```

Similarly, for peptide-level data:

```{r plot_pca_peptides, fig.height=7, fig.show='hold', fig.width=8, message=FALSE, warning=FALSE}
plot_PCA(
    pbf,
    pbf_name = c(
        "peptide::log2_on_raw",
        "peptide::medianNorm_on_log2_on_raw",
        "peptide::limmaRBE_on_medianNorm_on_log2_on_raw",
        "peptide::combat_on_medianNorm_on_log2_on_raw"
    ),
    sample_id_col = "Run",
    color_scheme = color_scheme,
    color_by = "Lab",
    shape_by = "Condition",
    fill_the_missing = NULL,
    base_size = 8, point_size = 3, point_alpha = 0.5
)
```

### t-SNE and UMAP

Core also provides `plot_TSNE()` and `plot_UMAP()` for matrices and
`ProBatchFeatures` objects. Like PCA, these functions treat matrix
columns as samples and align `colData()` by `sample_id_col`. Their
plotting-family `fill_the_missing` control is deliberately different
from the correction policy above: the default numeric value `-1` fills
missing values, while `NULL` (or `FALSE`) removes incomplete feature
rows. Both choices emit a warning when missing values are present.

A single selected assay returns a static `ggplot`. Supply
`random_state` explicitly when reproducible UMAP coordinates are
required; omitting it leaves randomness under the caller's control:

```{r plot_umap_single, fig.height=5, fig.width=6, eval=requireNamespace("umap", quietly=TRUE)}
umap_protein <- plot_UMAP(
    pbf,
    pbf_name = "protein::combat_on_medianNorm_on_log2_on_raw",
    sample_id_col = "Run",
    color_by = "Lab",
    shape_by = "Condition",
    color_scheme = color_scheme,
    fill_the_missing = NULL,
    n_neighbors = 10,
    random_state = 2026,
    plot_title = "UMAP after protein-level ComBat correction"
)
umap_protein
```

For multiple assays, `pbf_name = NULL` selects assays in object order,
whereas an explicit vector preserves the requested order. Shared
arguments are supplied normally, and `assay_args` is a named list keyed
by selected assay for settings that differ between assays. The next
example requests a specific order, assigns per-assay t-SNE seeds and
backend settings, and asks for both the arranged grob and its named plot
list:

```{r plot_tsne_multi, fig.height=5, fig.width=9, eval=requireNamespace("Rtsne", quietly=TRUE) && requireNamespace("gridExtra", quietly=TRUE)}
embedding_assays <- c(
    "protein::limmaRBE_on_medianNorm_on_log2_on_raw",
    "protein::combat_on_medianNorm_on_log2_on_raw"
)
n_embedding_samples <- ncol(pb_assay_matrix(pbf, embedding_assays[[1L]]))
tsne_perplexity <- min(
    10,
    max(1, floor((n_embedding_samples - 1) / 3))
)
tsne_args <- setNames(
    list(
        list(
            perplexity = tsne_perplexity,
            max_iter = 300,
            random_seed = 2026
        ),
        list(
            perplexity = tsne_perplexity,
            max_iter = 300,
            random_seed = 2027
        )
    ),
    embedding_assays
)

tsne_protein <- plot_TSNE(
    pbf,
    pbf_name = embedding_assays,
    sample_id_col = "Run",
    color_by = "Lab",
    shape_by = "Condition",
    color_scheme = color_scheme,
    fill_the_missing = NULL,
    plot_title = setNames(
        c("limma corrected", "ComBat corrected"),
        embedding_assays
    ),
    assay_args = tsne_args,
    return_gridExtra = TRUE,
    plot_ncol = 2
)
names(tsne_protein$plots) # same order as embedding_assays
grid::grid.draw(tsne_protein$grob)
```

`random_seed` controls t-SNE reproducibility. If a requested perplexity
is too large for an assay's sample count, Core reduces it to the
supported maximum and warns.

Static rendering is the default and does not require Plotly. With
`use_plotlyrender = TRUE`, one assay returns one `plotly` object and
multiple assays return a named ordered list. Set
`return_subplots = TRUE` to combine a multi-assay result; `subplot_ncol`
and `share_axes` control its layout. This optional interactive variant
is shown without executing it during vignette builds:

```{r plot_umap_interactive, eval=FALSE}
embedding_assays <- c(
    "protein::limmaRBE_on_medianNorm_on_log2_on_raw",
    "protein::combat_on_medianNorm_on_log2_on_raw"
)
umap_args <- setNames(
    list(
        list(n_neighbors = 10, random_state = 2026),
        list(n_neighbors = 15, random_state = 2027)
    ),
    embedding_assays
)

umap_plotly_list <- plot_UMAP(
    pbf,
    pbf_name = embedding_assays,
    sample_id_col = "Run",
    color_by = "Lab",
    shape_by = "Condition",
    color_scheme = color_scheme,
    fill_the_missing = NULL,
    assay_args = umap_args,
    use_plotlyrender = TRUE
)
names(umap_plotly_list) # same order as embedding_assays

umap_subplot <- plot_UMAP(
    pbf,
    pbf_name = embedding_assays,
    sample_id_col = "Run",
    color_by = "Lab",
    shape_by = "Condition",
    color_scheme = color_scheme,
    fill_the_missing = NULL,
    assay_args = umap_args,
    use_plotlyrender = TRUE,
    return_subplots = TRUE,
    subplot_ncol = 2,
    share_axes = c(x = TRUE, y = TRUE)
)
umap_subplot
```

### Hierarchical clustering

Hierarchical clustering provides another way to explore sample similarities.
Here we display clustering results for
`protein::combat_on_medianNorm_on_log2_on_raw` and
`protein::limmaRBE_on_medianNorm_on_log2_on_raw` assays and colour
samples by lab and condition:

```{r plot_hclust_combat, fig.show='hold', fig.width=12, fig.height=4}
plot_hierarchical_clustering(
    pbf,
    pbf_name = "protein::combat_on_medianNorm_on_log2_on_raw",
    sample_id_col = "Run",
    label_font = 0.6,
    color_list = color_scheme
)
```

```{r plot_hclust_limma, fig.show='hold', fig.width=12, fig.height=4}
plot_hierarchical_clustering(
    pbf,
    pbf_name = "protein::limmaRBE_on_medianNorm_on_log2_on_raw",
    sample_id_col = "Run",
    label_font = 0.6,
    color_list = color_scheme,
    fill_the_missing = NULL
)
```

### Principal Variance Component Analysis (PVCA)

PVCA quantifies the contribution of known factors to the observed
variance [@bushel2025pvca] which can be compared across raw, normalised, and
ComBat-corrected assays:

```{r plot_pvca_panels, fig.height=6, fig.show='hold', fig.width=5, message=FALSE, warning=FALSE}
plot_PVCA(
    pbf,
    pbf_name = c(
        "protein::medianNorm_on_log2_on_raw",
        "protein::combat_on_medianNorm_on_log2_on_raw"
    ),
    sample_id_col = "Run",
    technical_factors = c("Lab"),
    biological_factors = c("Condition"),
    fill_the_missing = NULL,
    base_size = 7,
    plot_ncol = 2, # the number of plots in a row
    variance_threshold = 0 # the percentile value of weight each of the
    # covariates needs to explain
    # (the rest will be lumped together)
)
```


# Inspecting operation log

Each call to `pb_transform()` or `pb_filterNA()` creates a log entry that
records the source assay, the applied operation, and the resulting assay name.
Reviewing this log ensures that complex processing pipelines remain
transparent and reproducible.

```{r show_operation_log}
get_operation_log(pbf) %>%
    as_tibble()
```

If you need a compact name for the current assay chain (for example, to
label benchmark results), use
`pb_pipeline_name(pbf, "protein::combat_on_medianNorm_on_log2_on_raw")`.

```{r compute_pipeline_name}
pb_pipeline_name(pbf, "protein::combat_on_medianNorm_on_log2_on_raw")
```

## Extracting data matrices from `pbf` object

It is possible to extract a specific data matrix from a `ProBatchFeatures`
object using the assay name:

```{r matr}
extracted_matrix <- pb_assay_matrix(
    pbf,
    assay = "peptide::raw"
)
# show
extracted_matrix[1:5, 1:5]
rm(extracted_matrix)
```

For consistency with the main **ProBatch** vignette, assay can also be
extracted in the long format. In this case, the
`sample_id_col` needs to be specified:

```{r extract_to_long}
extracted_long <- pb_as_long(
    pbf,
    sample_id_col = "Run",
    pbf_name = "peptide::raw"
)

# show
extracted_long[1:3, ]
rm(extracted_long)
```

# Session info

```{r sessionInfo}
sessionInfo()
```
# Citation

To cite this package, please use:

```{r citation}
citation("proBatch")
```

# References
