---
title: "Using PRONE Normalization in proBatch Pipelines"
author: |
  | Yuliya Burankova
  | Institute for Computational Systems Biology, University of Hamburg, Germany
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
package: proBatch
vignette: >
  %\VignetteIndexEntry{Using PRONE normalization methods in proBatch pipelines}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

This vignette shows a simple, practical workflow:

1. list PRONE normalization steps available in `pb_transform()`,
2. run one PRONE normalization step,
3. chain PRONE normalization with a batch-correction step.

```{r setup, message=FALSE}
library(proBatch)

prone_ok <- requireNamespace("PRONE", quietly = TRUE)
if (!prone_ok) {
    cat(
        "Package 'PRONE' is not installed. ",
        "PRONE-dependent chunks are shown but not evaluated.\n",
        sep = ""
    )
}
```

# Prepare example data

```{r load-data, eval=prone_ok}
data("tuberculosis_TMT_se", package = "PRONE")
se_src <- tuberculosis_TMT_se

stopifnot("raw" %in% SummarizedExperiment::assayNames(se_src))
mat <- as.matrix(SummarizedExperiment::assay(se_src, "raw"))
sample_ann <- as.data.frame(SummarizedExperiment::colData(se_src))

sample_id_col <- "Column"
stopifnot(sample_id_col %in% colnames(sample_ann))
sample_ann <- sample_ann[match(colnames(mat), as.character(sample_ann[[sample_id_col]])), , drop = FALSE]
stopifnot(identical(as.character(sample_ann[[sample_id_col]]), colnames(mat)))

md <- S4Vectors::metadata(se_src)
batch_col <- md[["batch"]]
if (!is.character(batch_col) || length(batch_col) != 1L ||
    !(batch_col %in% colnames(sample_ann))) {
    batch_col <- NULL
}

.prone_step_params_for <- function(step_name) {
    step_token <- tolower(sub("^pr", "", step_name, ignore.case = TRUE))
    needs_batch <- step_token %in% c("irs", "limbe")
    needs_refs <- step_token %in% c("irs")
    needs_condition <- step_token %in% c("eigenms")

    params <- list(sample_id_col = sample_id_col)

    if (needs_batch) {
        if (is.null(batch_col)) {
            stop(
                "Step '", step_name,
                "' requires PRONE `batch`, but no valid batch column is available."
            )
        }
        params$batch <- batch_col
    }

    if (needs_refs) {
        refs_val <- md[["refs"]]
        if (is.null(refs_val)) {
            stop(
                "Step '", step_name,
                "' requires PRONE `refs`, but metadata(se)$refs is missing."
            )
        }
        params$refs <- refs_val
    }

    if (needs_condition) {
        condition_val <- md[["condition"]]
        if (is.null(condition_val)) {
            stop(
                "Step '", step_name,
                "' requires PRONE `condition`, but metadata(se)$condition is missing."
            )
        }
        params$condition <- condition_val
    }

    Filter(Negate(is.null), params)
}

pbf <- ProBatchFeatures(
    data_matrix = mat,
    sample_annotation = sample_ann,
    sample_id_col = sample_id_col,
    name = "raw"
)

pbf
```

# Find PRONE steps for `pb_transform()`

```{r register-prone-steps, eval=prone_ok}
# Methods available in PRONE itself
head(PRONE::get_normalization_methods())

# PRONE normalization steps currently available in proBatch
prone_steps <- pb_list_steps("^PR")
if (!length(prone_steps)) {
    stop("No PRONE normalization steps are currently available in pb_list_steps().")
}
head(prone_steps, 10)

# batch-correction steps available in the current session
intersect(c("combat", "limmaRBE", "mComBat", "loessLimmaRBE"), pb_list_steps())
```

PRONE normalization methods are registered automatically at package load
when `PRONE` is installed. If a method name collides with an existing
`proBatch` step (for example `medianNorm`), the original `proBatch`
implementation keeps the unprefixed name, and the PRONE version remains
available via `PR<method>`.

# Choose one PRONE step for the example

```{r choose-prone-step, eval=prone_ok}
preferred_step <- "PRGlobalMedian"
chosen_step <- if (preferred_step %in% prone_steps) preferred_step else prone_steps[[1]]
chosen_step
```

# Apply one PRONE normalization step with `pb_transform()`

```{r prone-transform-only, eval=prone_ok}
chosen_step_params <- .prone_step_params_for(chosen_step)

pbf_norm <- pb_transform(
    object = pbf,
    from = "feature::raw",
    steps = chosen_step,
    params_list = list(chosen_step_params)
)

# New assay created by this step
tail(names(pbf_norm), 1)
```

# Chain PRONE normalization with one BEC step in `pb_transform()`

```{r prone-transform-with-bec, eval=prone_ok}
available_bec <- pb_list_steps()
bec_candidates <- intersect(c("combat", "limmaRBE"), available_bec)
if (!length(bec_candidates)) {
    stop("Neither 'combat' nor 'limmaRBE' is currently available.")
}
bec_step <- bec_candidates[[1]]
if (is.null(batch_col)) {
    stop("Batch-correction step '", bec_step, "' requires a valid batch column.")
}
bec_params <- switch(bec_step,
    combat = list(
        batch_col = batch_col,
        sample_id_col = sample_id_col,
        par.prior = TRUE,
        fill_the_missing = 0
    ),
    limmaRBE = list(
        batch_col = batch_col,
        sample_id_col = sample_id_col
    )
)

pbf_norm_bec <- pb_transform(
    object = pbf,
    from = "feature::raw",
    steps = c(chosen_step, bec_step),
    params_list = list(
        .prone_step_params_for(chosen_step),
        bec_params
    )
)

# Show assays created by this chain
grep("^feature::(PR|combat|limmaRBE)", names(pbf_norm_bec), value = TRUE)

# inspect operation log entries for these pipelines
oplog_df <- as.data.frame(get_operation_log(pbf_norm_bec))
tail(oplog_df[, c("step", "from", "to")], 4)
```

# Optional: evaluate a chain quickly without storing assays

```{r prone-pbeval-quick, eval=prone_ok}
mat_eval <- pb_eval(
    object = pbf,
    from = "feature::raw",
    steps = c(chosen_step, bec_step),
    params_list = list(
        .prone_step_params_for(chosen_step),
        bec_params
    )
)

dim(mat_eval)
summary(as.vector(mat_eval))
```

# Optional: smoke-test all PRONE normalization steps

```{r prone-all-methods-smoke, eval=prone_ok}
target_methods <- PRONE::get_normalization_methods()
target_steps <- unique(c(
    paste0("PR", target_methods),
    tolower(paste0("PR", target_methods))
))
target_steps <- intersect(target_steps, pb_list_steps("^PR|^pr"))

smoke_results <- do.call(
    rbind,
    lapply(target_steps, function(step_name) {
        out <- try(
            {
                step_params <- .prone_step_params_for(step_name)
                pb_eval(
                    object = pbf,
                    from = "feature::raw",
                    steps = step_name,
                    params_list = list(step_params)
                )
            },
            silent = TRUE
        )
        ok <- !inherits(out, "try-error")
        msg <- if (ok) "" else as.character(out)
        data.frame(
            step = step_name,
            ok = ok,
            message = msg,
            stringsAsFactors = FALSE
        )
    })
)

smoke_results
subset(smoke_results, !ok)
```

# Notes

- PRONE normalization methods are registered as standard
  `pb_transform()`/`pb_eval()` steps, so they can be mixed with any other
  registered proBatch step.
- In this vignette, step parameters are built with
  `.prone_step_params_for(step_name)`, which adds PRONE metadata only when
  needed:
  `batch` for `limBE`/`IRS`, `refs` for `IRS`, and `condition` for `EigenMS`.
- Use `pb_list_steps("^PR")` to list currently available PRONE
  normalization steps.
- For reproducible benchmarking, store pipeline outputs with
  `pb_transform()` and compare diagnostics across assays.
